Hilbert space methods to approximate Gaussian processes using Stan

### 2019-04-18 Paper/manuscript/manuscript_05.tex
improve section 2.2

% interacttfssample.tex
% v1.05 - August 2017
\documentclass[]{interact}
\usepackage[utf8]{inputenc}
\usepackage{times}
\usepackage{microtype}

@@

@@

In this work, Gaussian processes will take the role of a prior distribution over function spaces for non-parametric latent functions in a Bayesian setting. 
Consider a data set $\mathcal{D} = \left\lbrace \mathbf{x}_n, y_n \right\rbrace_{n=1}^N$, where $y_n$ is modelled conditionally as $p(y_n|f(\mathbf{x}_n),\phi)$, where $p$ is some parametric distribution with parameters $f$ and $\phi$, and $f$ is an unknown function with Gaussian process prior. This generalizes trivially to more complex models depending on several unknown functions, for example such as $p(y_n|f(\mathbf{x}_n),g(\mathbf{x}_n))$ or multilevel models. Our goal is to obtain posterior distribution for the value of the function $f^*=f(\mathbf{x}^*)$  evaluated at a new input point $\mathbf{x}^* \in \mathbb{R}^D$.
% That is, we want to obtain the predictive distribution $p(f^*|\mathcal{D})$ of $f^*$ conditioned on the data $\mathcal{D}$.\\

We assume a Gaussian process prior for $f \sim \mathcal{GP}(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x}'))$, where $\mu: \mathbb{R}^D \rightarrow \mathbb{R}$ and $k: \mathbb{R}^D \times \mathbb{R}^D \rightarrow \mathbb{R}$ are the mean and covariance functions, respectively,
%
\begin{align*}
 	\mu(\mathbf{x}) &= \mathbb{E}\!\left[f(\mathbf{x})\right],\\ 
 	k(\mathbf{x}, \mathbf{x}') &= \mathbb{E}\!\left[\left( f(\mathbf{x}) - \mu(\mathbf{x}) \right)\left( f(\mathbf{x}') - \mu(\mathbf{x}') \right)\right].
\end{align*}

The mean and covariance functions completely characterize the Gaussian process prior, and control the a priori behavior of the function $f$. Let $\mathbf{f}=\left\lbrace f(\mathbf{x}_n) \right\rbrace_{n=1}^N$, then the resulting prior distribution for $\mathbf{f}$ is a multivariate Gaussian distribution
$\mathbf{f} \sim \mathcal{N}(\boldsymbol{\mu}, \mathbf{K})$,
 where $\boldsymbol{\mu} = \left\lbrace \mu(\mathbf{x}_n) \right\rbrace_{n=1}^N$ is the mean and $\mathbf{K}$ the covariance matrix, where $K_{i,j}=k(\mathbf{x}_i,\mathbf{x}_j)$. The covariance function $k(\textbf{x}, \textbf{x}')$ might depend on a set of hyperparameters, $\bm{\theta}$, but we will not write this dependency explicitly to ease the notation.
The joint distribution of $\mathbf{f}$ and a new $f^*$ is also a multivariate Gaussian as,
%
\begin{align*}
p(\mathbf{f}, f^*)=\mathcal{N} \left( \left[ \begin{array}{cc}
\mathbf{f} \\ 
f^*
\end{array} \right] \,\middle|\, \mathbf{0},\left[ \begin{array}{cc}
\mathbf{K}_{\mathbf{f},\mathbf{f}} & \mathbf{k}_{\mathbf{f},f^*} \\ 
\mathbf{k}_{f^*,\mathbf{f}} & k_{f^*,f^*}
\end{array} \right] \right),
\end{align*}

\noindent where $\mathbf{k}_{\mathbf{f},f^*}$ is the covariance between $\mathbf{f}$ and $f^*$, and $k_{f^*,f^*}$ is the prior variance of $f^*$. %By using the conditioning properties of multivariate Gaussian distributions, we can derive the predictive distribution for $f^*$  given $\mathbf{f}$ analytically,
% \begin{align*}
	% p(f^*| \mathbf{f}) = \mathcal{N}(f^*| \mathbf{k}_{f^*,\mathbf{f}} \mathbf{K}_{\mathbf{f},\mathbf{f}}^{-1} \mathbf{f},  k_{f^*,f^*}-\mathbf{k}_{f^*,\mathbf{f}} \mathbf{K}_{\mathbf{f},\mathbf{f}}^{-1} \mathbf{k}_{\mathbf{f},f^*}  )
% \end{align*}

@@

% The joint distribution of the observations $\mathbf{y} = \left\lbrace y_n \right\rbrace_{n=1}^N$ and function values $\mathbf{f}$ and $f^*$, $p(\mathbf{y}, \mathbf{f}, f^*)$, is the product of the conditional distribution for $\mathbf{y}$ given $\mathbf{f}$ and the joint distribution for $\mathbf{f}$ and $f^*$,
% \begin{align*}
% 	p(\mathbf{y}, \mathbf{f}, f^*) = p(\mathbf{y}|\mathbf{f})p(\mathbf{f}, f^*)
% \end{align*}

% By marginalizing over $\mathbf{f}$ and conditioning on vector of observations $\mathbf{y}$, we obtain the posterior distribution of interest
% %
% \begin{align}\label{eq:posterior}
% 	p(f^*|\mathbf{y}) = \frac{\int p(\mathbf{y}|\mathbf{f})p(\mathbf{f}, f^*)\text{d}\mathbf{f}}{p(\mathbf{y})},
% \end{align}
% where $p(\mathbf{y})$ is the marginal likelihood and is given by
% %
% \begin{align}\label{eq:evidence}
% 	p(\textbf{y}) = \int p(\mathbf{y}|\mathbf{f})p(\mathbf{f}, f^*) \text{d}\bm{f}\text{d}f^*.
% \end{align}
% %
% If the observation model $p(\mathbf{y}|\mathbf{f})$ is Gaussian, both integrals in eq. \eqref{eq:posterior} and \eqref{eq:evidence} can be solved analytically conditioned on the hyperparameters. For example, an isotropic Gaussian likelihood yields the following closed-form solution
% \begin{align*}
% 	p(f^*| \mathbf{y}) = \mathcal{N}(f^*| \textbf{k}_{f^*,\mathbf{f}} (\textbf{K}_{\mathbf{f},\mathbf{f}} + \sigma^2 \mathbf{I})^{-1} \mathbf{f},  k_{f^*,f^*}-\mathbf{k}_{f^*,\mathbf{f}} (\mathbf{K}_{\mathbf{f},\mathbf{f}} + \sigma^2 \mathbf{I})^{-1} \mathbf{k}_{\mathbf{f},f^*}  ),
% \end{align*}
% where $\sigma^2$ is the noise variance.

If $p(y_n|f(\mathbf{x}_n),\phi)=N(y_n|f(\mathbf{x}_n),\sigma)$ then $\mathbf{f}$ can be integrated out analytically (with a computational cost of $O(n^3)$ for exact GP and $O(nm^2)$ for sparse GP). If $p(y_n|f(\mathbf{x}_n),\phi)=N(y_n|f(\mathbf{x}_n),g(\mathbf{x}_n))$ or $p(y_n|f(\mathbf{x}_n),\phi)$ is non-Gaussian, the marginalization does not have closed form solution.
%
Furthermore, if a prior distribution is imposed on $\phi$ and $\theta$ to form a joint posterior for $\phi$, $\theta$ and $\mathbf{f}$,
approximate inference such as Markov chain Monte Carlo (MCMC) \citep{brooks_2011} Laplace approximation (\citep{williams1998bayesian,rasmussen2006gaussian}, expectation propagation \citep{minka2001expectation}, or variational Bayes methods \citep{gibbs2000variational,csato2000efficient} need to be used.
%
In this paper we focus on use of MCMC for integrating over the joint
posterior. MCMC is not usually the fastest approach, but allows
accurate inference for general models in probabilistic programming
setting. We consider the computational costs of GPs specifically from
this point of view.

\subsection{Covariance function}

The covariance function is the crucial ingredient in a Gaussian process as it encodes our prior assumptions about the function, and defines a correlation structure which characterize the correlations between function values at different inputs. A stationary covariance function is a function of $\bm{\tau}=\mathbf{x}-\mathbf{x}' \in \mathbb{R}^D$, such that it can be written $k(\mathbf{x},\mathbf{x}') = k(\bm{\tau})$, which means that the covariance is invariant to translations. Isotropic covariance functions are those that are function of the distance between observations, $k(\mathbf{x},\mathbf{x}') = k(||\mathbf{x}-\mathbf{x}'||) = k(r), r\in \mathbb{R}$, which means that the covariance is both translation and rotation invariant. The Mat\'ern family with half-integers is the class of isotropic covariance functions probably most commonly used, which, in the unidimensional case (i.e. $\mathbf{x}=x\in \mathbb{R}$, so $\mathbf{r}=r\in \mathbb{R}$, and $\ell\in \mathbb{R}$ ), are
%
\begin{align*}
k_{\nu=\infty}(r)&=\sigma^2 \exp(-\frac{1}{2} r^2/\ell^2),  \\
k_{\nu=\frac{1}{2}}(r)&=\sigma^2 \exp(-r/\ell),  \\
k_{\nu=\frac{3}{2}}(r)&=\sigma^2(1+\sqrt{3}r/\ell) \exp(-\sqrt{3}r/\ell), 
%k_{\nu=\frac{5}{2}}(r)&=\sigma^2(1+\sqrt{5}r/\ell+\frac{5}{3}r^2/\ell^2) \exp(-\sqrt{5}r/\ell), 
\end{align*}

\noindent where $\nu$ is the order the kernel, and the $\ell$ and $\sigma$ are the length-scale and magnitude, respectively, of the kernel. The particular case where $\nu=\infty$ is commonly known as squared exponential (exponentiated quadratic) covariance function, and that with $\nu=1/2$ as exponential covariance function. These covariance functions can be easily generalized to a $D$-dimensional case, by using the property of separability of multidimensional kernels. Let $\mathbf{x}\in \mathbb{R}^D$, $\mathbf{\bm{\tau}}=\{\tau_1,\tau_2\}\in \mathbb{R}^D$ and $\boldsymbol{\ell}=\{\ell_1,\ell_2\}\in \mathbb{R}^D$,
%
\begin{align*}
k_{\nu=\infty}(\bm{\tau})&=\exp(-\frac{1}{2} \tau_1^2/\ell_1^2) \cdot \exp(-\frac{1}{2} \tau_2^2/\ell_2^2), \\
k_{\nu=\frac{1}{2}}(\bm{\tau})&=\exp(-\tau_1/\ell_1) \cdot \exp(-\tau_2/\ell_2), \\
k_{\nu=\frac{3}{2}}(\bm{\tau})&=(1+\sqrt{3}\tau_1/\ell_1) \exp(-\sqrt{3}\tau_1/\ell_1) \cdot (1+\sqrt{3}\tau_2/\ell_2) \exp(-\sqrt{3}\tau_2/\ell_2) \\
\end{align*}

\noindent Using a multidimensional length-scale basically turns the isotropic covariance function into non-isotropic.

Stationary covariance functions can be represented in terms of their spectral densities \citep{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure (\textit{Bochner's theorem}, see, e.g. \cite{akhiezer1993theory}). If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. In the unidimensional case the spectral density functions associated with the Mat\'ern covariance functions written above are
%
\begin{align}
S_{\nu=\infty}(w)&= \sigma^2 \sqrt{2\pi} \cdot \ell \cdot \exp\left(-0.5 \ell^2 w^2\right), \label{eq:specdens_inf}  \\
S_{\nu=\frac{1}{2}}(w)&= 2\sigma^2 \frac{1}{\ell}\left(\frac{1}{\ell^2} + \omega^2\right)^{-1}, \label{eq:specdens_12}  \\
S_{\nu=\frac{3}{2}}(w)&= 4\sigma^2 \frac{\sqrt{3}}{\ell}^{3}\left(\frac{3}{\ell^2} + \omega^2\right)^{-2}, \label{eq:specdens_32} \\
%S_{\nu=\frac{5}{2}}(w)&= \frac{16}{3}\sigma^2 \frac{\sqrt{3}}{\ell}^{5}\left(\frac{5}{\ell^2} + \omega^2\right)^{-3}, \label{eq:specdens_52} 
\end{align}

\noindent where variable $w\in \mathbb{R}$ is in the frequency domain, and $\ell\in \mathbb{R}$. By using the separability property of the spectral densities functions, the $D$-dimensional form of these spectral density functions, where $\mathbf{w}=\{w_1,w_2\}$ and $\mathbf{\ell}=\{\ell_1,\ell_2\}$ are
%
\begin{align}
S_{\nu=\infty}(\bm{\omega})&= \sqrt{2\pi} \cdot \ell_1 \cdot \exp\left(-0.5 \ell_1^2 \omega_1^2\right) \cdot  \sqrt{2\pi} \cdot \ell_2 \cdot \exp\left(-0.5 \ell_2^2 \omega_2^2\right) \label{eq:specdens_multi_inf} \\
S_{\nu=\frac{1}{2}}(\bm{\omega})&= 2 \frac{1}{\ell_1}\left(\frac{1}{\ell_1^2} + \omega_1^2\right)^{-1} \cdot 2 \frac{1}{\ell_2}\left(\frac{1}{\ell_2^2} + \omega_2^2\right)^{-1} \label{eq:specdens_multi_12} \\
S_{\nu=\frac{3}{2}}(\bm{\omega})&= 4 \frac{\sqrt{3}}{\ell_1}^{3}\left(\frac{3}{\ell_1^2} + \omega_1^2\right)^{-2} \cdot 4 \frac{\sqrt{3}}{\ell_2}^{3}\left(\frac{3}{\ell_2^2} + \omega_2^2\right)^{-2}, \label{eq:specdens_multi_32}
\end{align}

\noindent which can also be derived from the following expressions \citep{rasmussen2006gaussian}
%
\begin{align*}
S^{\ast}_{\nu=\infty}(\mathbf{\omega})&= \sigma^2 \sqrt{2\pi}^D \prod_{d=1}^D l_d \cdot \exp\left(-0.5 \sum_{d=1}^D l_d^2 \mathbf{\omega}_d^2\right), \\
%
S^{\ast}_{\nu}(\mathbf{\omega})&= \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)l^{2\nu}}\left(\frac{2\nu}{l^2}+4\pi^2\mathbf{\omega}^2 \right)^{(\nu+D/2)}.
\end{align*}

@@

If we truncate the sum in (\ref{approxcov}) to the first $m$ terms, the approximate covariance function becomes
%
\begin{equation}
k(x,x') \approx \sum_{j=1}^m S_{\theta}(\sqrt{\lambda_j}) \phi_j(x) \phi_j(x') = \boldsymbol{\phi}(x)^\intercal \Delta \boldsymbol{\phi}(x'), \nonumber
\end{equation}

\noindent where $\boldsymbol{\phi}(x)=\{\phi_j(x)\}_{j=1}^{m} \in {\rm I\!R}^{m}$ is the column vector of basis functions, and $\Delta  \in {\rm I\!R}^{m\times m}$ is the diagonal matrix of the spectral densities $S_{\theta}(\sqrt{\lambda_j})$: 
%
\begin{eqnarray}
\Delta &=&  \begin{bmatrix}
    S_{\theta}(\sqrt{\lambda_1}) & & \\
    & \ddots & \nonumber \\
    & & S_{\theta}(\sqrt{\lambda_m}) \\
  \end{bmatrix}.
\end{eqnarray}

@@

\noindent As a result, the model for $f$ can be written as
%
\begin{equation}
\mathbf{f} \sim \mathcal{N}(\boldsymbol{\mu},\Phi \Delta \Phi^\intercal). \nonumber
\end{equation}

@@

Each multivariate eigenfunction $\phi^{\ast}_j$ corresponds to the product of the univariate eigenfunctions whose indices corresponds to the elements of the $D$-tuple $\mathbb{S}_{j\cdotp}$, and each multivariate eigenvalue $\boldsymbol{\lambda}^{\ast}_j$ is a $D$-vector with elements that are the univariate eigenvalues whose indices correspond to the elements of the $D$-tuple $\mathbb{S}_{j\boldsymbol{\cdotp}}$. Thus, for $\mathbf{x}=\{x_d\}_{d=1}^D \in \Omega$ and $j=1,\ldots,m^{\ast}$, we have 
%
\begin{align}
\phi^{\ast}_j(\mathbf{x}) &= \prod_{d=1}^{D} \phi_{\mathbb{S}_{jd}}(x_d) = \prod_{d=1}^{D} \sqrt{\frac{1}{L_d}} \text{sin}\left(\sqrt{\lambda_{\mathbb{S}_{jd}}}(x_d+L_d)\right) \label{eigenvalue_multi} \\
%
\boldsymbol{\lambda}^{\ast}_j &= \left\{ \lambda_{\mathbb{S}_{jd}} \right\}_{d=1}^D =  \left\{ \left(\tfrac{\pi \mathbb{S}_{jd}}{2L_d}\right)^2 \right\}_{d=1}^D. \label{eigenfunction_multi}
\end{align}

\noindent The approximate covariance function is then represented as
%
\begin{equation}\label{approxcov_multi}
k(\mathbf{x},\mathbf{x}') \approx \sum_{j=1}^{m^{\ast}} 
S^{\ast}_{\theta}\left(\sqrt{\boldsymbol{\lambda}^{\ast}_j}\right)
\phi^{\ast}_j(\mathbf{x}) \phi^{\ast}_j(\mathbf{x}'),
\end{equation}

\noindent where $S^{\ast}_{\theta}$ is the spectral density of the $D$-dimensional covariance function (see Section 2.1). We can now write the approximate series expansion of the multivariate function $f$ as,
%
\begin{equation}\label{approxf_multi}
f(\mathbf{x}) \approx \sum_{j=1}^{m^{\ast}} 
\left( S^{\ast}_{\theta} \left(\sqrt{\boldsymbol{\lambda}^{\ast}_j} \right)\right)^{1/2}
\phi^{\ast}_j(\mathbf{x}) \beta_j, 
\end{equation}

@@

This example consists of a simulated dataset with 250 ($i=1,\dots,n=250$) single draws of a Gaussian process prior with a Mat{\'e}rn($\nu$=3/2) covariance function and hyperparameters marginal variance $\alpha=1$ and lengthscale $\ell=0.15$, with corresponding inputs values $\mathbf{x}=\{x_i\}_{i=1}^{250}$ with $x_i \in [-1,1] \subset {\rm I\!R}$. To form the final noisy dataset $\mathbf{y}$, Gaussian noise $\sigma=0.2$ was added to the GP draws.

>>>>

% interacttfssample.tex
% v1.05 - August 2017
\documentclass[]{interact}
\usepackage[utf8]{inputenc}
\usepackage{times}
\usepackage{microtype}
\usepackage{lscape}

@@

\newenvironment{changemargin}[2]{%
\begin{list}{}{%
\setlength{\topsep}{0pt}%
\setlength{\leftmargin}{#1}%
\setlength{\rightmargin}{#2}%
\setlength{\listparindent}{\parindent}%
\setlength{\itemindent}{\parindent}%
\setlength{\parsep}{\parskip}%
}%
\item[]}{\end{list}}

@@

In this work, Gaussian processes will take the role of a prior distribution over function spaces for non-parametric latent functions in a Bayesian setting. 
Consider a data set $\mathcal{D} = \left\lbrace \bm{x}_n, y_n \right\rbrace_{n=1}^N$, where $y_n$ is modelled conditionally as $p(y_n|f(\bm{x}_n),\phi)$, where $p$ is some parametric distribution with parameters $f$ and $\phi$, and $f$ is an unknown function with Gaussian process prior. This generalizes trivially to more complex models depending on several unknown functions, for example such as $p(y_n|f(\bm{x}_n),g(\bm{x}_n))$ or multilevel models. Our goal is to obtain posterior distribution for the value of the function $f^*=f(\bm{x}^*)$  evaluated at a new input point $\bm{x}^* \in \mathbb{R}^D$.
% That is, we want to obtain the predictive distribution $p(f^*|\mathcal{D})$ of $f^*$ conditioned on the data $\mathcal{D}$.\\

We assume a Gaussian process prior for $f \sim \mathcal{GP}(\mu(\bm{x}), k(\bm{x}, \bm{x}'))$, where $\mu: \mathbb{R}^D \rightarrow \mathbb{R}$ and $k: \mathbb{R}^D \times \mathbb{R}^D \rightarrow \mathbb{R}$ are the mean and covariance functions, respectively,
%
\begin{align*}
 	\mu(\bm{x}) &= \mathbb{E}\!\left[f(\bm{x})\right],\\ 
 	k(\bm{x}, \bm{x}') &= \mathbb{E}\!\left[\left( f(\bm{x}) - \mu(\bm{x}) \right)\left( f(\bm{x}') - \mu(\bm{x}') \right)\right].
\end{align*}

The mean and covariance functions completely characterize the Gaussian process prior, and control the a priori behavior of the function $f$. Let $\bm{f}=\left\lbrace f(\bm{x}_n) \right\rbrace_{n=1}^N$, then the resulting prior distribution for $\bm{f}$ is a multivariate Gaussian distribution
$\bm{f} \sim \mathcal{N}(\bm{\mu}, \bm{K})$,
 where $\bm{\mu} = \left\lbrace \mu(\bm{x}_n) \right\rbrace_{n=1}^N$ is the mean and $\bm{K}$ the covariance matrix, where $K_{i,j}=k(\bm{x}_i,\bm{x}_j)$. The covariance function $k(\bm{x}, \bm{x}')$ might depend on a set of hyperparameters, $\bm{\theta}$, but we will not write this dependency explicitly to ease the notation.
The joint distribution of $\bm{f}$ and a new $f^*$ is also a multivariate Gaussian as,
%
\begin{align*}
p(\bm{f}, f^*)=\mathcal{N} \left( \left[ \begin{array}{cc}
\bm{f} \\ 
f^*
\end{array} \right] \,\middle|\, \bm{0},\left[ \begin{array}{cc}
\bm{K}_{\bm{f},\bm{f}} & \bm{k}_{\bm{f},f^*} \\ 
\bm{k}_{f^*,\bm{f}} & k_{f^*,f^*}
\end{array} \right] \right),
\end{align*}

\noindent where $\bm{k}_{\bm{f},f^*}$ is the covariance between $\bm{f}$ and $f^*$, and $k_{f^*,f^*}$ is the prior variance of $f^*$. %By using the conditioning properties of multivariate Gaussian distributions, we can derive the predictive distribution for $f^*$  given $\bm{f}$ analytically,
% \begin{align*}
	% p(f^*| \bm{f}) = \mathcal{N}(f^*| \bm{k}_{f^*,\bm{f}} \bm{K}_{\bm{f},\bm{f}}^{-1} \bm{f},  k_{f^*,f^*}-\bm{k}_{f^*,\bm{f}} \bm{K}_{\bm{f},\bm{f}}^{-1} \bm{k}_{\bm{f},f^*}  )
% \end{align*}

@@

% The joint distribution of the observations $\bm{y} = \left\lbrace y_n \right\rbrace_{n=1}^N$ and function values $\bm{f}$ and $f^*$, $p(\bm{y}, \bm{f}, f^*)$, is the product of the conditional distribution for $\bm{y}$ given $\bm{f}$ and the joint distribution for $\bm{f}$ and $f^*$,
% \begin{align*}
% 	p(\bm{y}, \bm{f}, f^*) = p(\bm{y}|\bm{f})p(\bm{f}, f^*)
% \end{align*}

% By marginalizing over $\bm{f}$ and conditioning on vector of observations $\bm{y}$, we obtain the posterior distribution of interest
% %
% \begin{align}\label{eq:posterior}
% 	p(f^*|\bm{y}) = \frac{\int p(\bm{y}|\bm{f})p(\bm{f}, f^*)\text{d}\bm{f}}{p(\bm{y})},
% \end{align}
% where $p(\bm{y})$ is the marginal likelihood and is given by
% %
% \begin{align}\label{eq:evidence}
% 	p(\textbf{y}) = \int p(\bm{y}|\bm{f})p(\bm{f}, f^*) \text{d}\bm{f}\text{d}f^*.
% \end{align}
% %
% If the observation model $p(\bm{y}|\bm{f})$ is Gaussian, both integrals in eq. \eqref{eq:posterior} and \eqref{eq:evidence} can be solved analytically conditioned on the hyperparameters. For example, an isotropic Gaussian likelihood yields the following closed-form solution
% \begin{align*}
% 	p(f^*| \bm{y}) = \mathcal{N}(f^*| \textbf{k}_{f^*,\bm{f}} (\textbf{K}_{\bm{f},\bm{f}} + \sigma^2 \bm{I})^{-1} \bm{f},  k_{f^*,f^*}-\bm{k}_{f^*,\bm{f}} (\bm{K}_{\bm{f},\bm{f}} + \sigma^2 \bm{I})^{-1} \bm{k}_{\bm{f},f^*}  ),
% \end{align*}
% where $\sigma^2$ is the noise variance.

If $p(y_n|f(\bm{x}_n),\phi)=N(y_n|f(\bm{x}_n),\sigma)$ then $\bm{f}$ can be integrated out analytically (with a computational cost of $O(n^3)$ for exact GP and $O(nm^2)$ for sparse GP). If $p(y_n|f(\bm{x}_n),\phi)=N(y_n|f(\bm{x}_n),g(\bm{x}_n))$ or $p(y_n|f(\bm{x}_n),\phi)$ is non-Gaussian, the marginalization does not have closed form solution.
%
Furthermore, if a prior distribution is imposed on $\phi$ and $\theta$ to form a joint posterior for $\phi$, $\theta$ and $\bm{f}$,
approximate inference such as Markov chain Monte Carlo (MCMC) \citep{brooks_2011} Laplace approximation (\citep{williams1998bayesian,rasmussen2006gaussian}, expectation propagation \citep{minka2001expectation}, or variational Bayes methods \citep{gibbs2000variational,csato2000efficient} need to be used.
%
In this paper we focus on use of MCMC for integrating over the joint
posterior. MCMC is not usually the fastest approach, but allows
accurate inference for general models in probabilistic programming
setting. We consider the computational costs of GPs specifically from
this point of view.

\subsection{Covariance function and spectral density}

The covariance function is the crucial ingredient in a Gaussian process as it encodes our prior assumptions about the function, and defines a correlation structure which characterize the correlations between function values at different inputs. A stationary covariance function is a function of $\bm{\tau}=\bm{x}-\bm{x}' \in {\rm I\!R}^D$, such that it can be written $k(\bm{x},\bm{x}') = k(\bm{\tau})$, which means that the covariance is invariant to translations. Isotropic covariance functions are those that are function of the distance between observations, $k(\bm{x},\bm{x}') = k(|\bm{x}-\bm{x}'|) = k(r), r\in {\rm I\!R}$, which means that the covariance is both translation and rotation invariant. The most commonly used distance between observations is the norm L2 $(|\bm{x}-\bm{x}'|_{L2})$, also known as Euclidean distance, although other types of distances can be considered.

The Mat\'ern class of isotropic covariance functions is given by, 
%which, in the unidimensional case (i.e. $\bm{x}=x\in \mathbb{R}$, so $\bm{r}=r\in \mathbb{R}$, and $\ell\in \mathbb{R}$ ), are
%
\begin{align*}
k_{\nu}(r)&=\frac{2^{1-\nu}}{\Gamma(\nu)}\left(\frac{\sqrt{2\nu}r}{\ell}\right)^{\nu}K_{\nu}\left(\frac{\sqrt{2\nu}r}{\ell}\right),
\end{align*}

\noindent where $\nu$ is the order the kernel, $K_{\nu}$ the modified Bessel function, and the $\ell$ and $\sigma$ are the length-scale and magnitude, respectively, of the kernel. The particular case where $\nu=\infty$ and $\nu=3/2$ are probably the most commonly used kernels \citep{rasmussen2006gaussian}, 
%
\begin{align*}
k_{\infty}(r)&=\sigma^2 \exp\bigg(-\frac{1}{2} \frac{r^2}{\ell^2}\bigg),  \\
k_{\frac{3}{2}}(r)&=\sigma^2\bigg(1+\frac{\sqrt{3}r}{\ell}\bigg) \exp\bigg(-\frac{\sqrt{3}r}{\ell}\bigg). 
\end{align*}

\noindent The former is commonly known as squared exponential (exponentiated quadratic) covariance function. Assuming the Euclidean distance between observations, $r=|\bm{x}-\bm{x}'|_{L2}=\sqrt{\sum_{i=1}^{D}(x_i-x_i')^2}$, the kernels written above take the form
%
\begin{align*}
k_{\infty}(|\bm{x}-\bm{x}'|_{L2})&=\exp\bigg(-\frac{1}{2} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\bigg),\\
k_{\frac{3}{2}}(|\bm{x}-\bm{x}'|_{L2})&=\Bigg(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\Bigg)\exp\Bigg(- \sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\Bigg). \\
\end{align*}

\noindent Notice that the previous expressions have been easily generalized to using a multidimensional length-scale $\bm{\ell}\in {\rm I\!R}^D$. The use of a multidimensional length-scale basically turns the isotropic covariance function into non-isotropic.

Stationary covariance functions can be represented in terms of their spectral densities \citep{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure (\textit{Bochner's theorem}, see, e.g. \cite{akhiezer1993theory}). If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. The spectral density functions associated with the Mat\'ern class of covariance functions is given by
%
\begin{align*}
S_{\nu}(\omega)&= \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)l^{2\nu}}\left(\frac{2\nu}{l^2}+4\pi^2\omega^2 \right)^{(\nu+D/2)}
\end{align*}

\noindent in $D$ dimensions, where variable $\omega\in {\rm I\!R}$ is a distance in the frequency domain, and $\ell$ and $\sigma$ are the lengthscale and magnitud, respectively, of the kernel. The particular cases where $\nu=\infty$ and $\nu=3/2$ take the form
%
\begin{align}
S_{\infty}(\omega)&= \sigma^2 \sqrt{2\pi}^D  \ell^D  \exp\left(-0.5 \ell^2 \omega^2\right), \label{eq:specdens_inf}  \\
S_{\frac{3}{2}}(\omega)&= \sigma^2 \frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2})\sqrt{3}^3}{\frac{1}{2}\sqrt{\pi}\ell^3}\left(\frac{3}{\ell^2}+\omega^2 \right)^{-\frac{D+3}{2}}. \label{eq:specdens_32} 
\end{align}

\noindent Particularizing to an input dimension $D=3$ and Euclidean distance $\omega=\sqrt{\sum_{i=1}^{D=3}s_i^2}$, and considering a multidimensional lengthscale $\bm{\ell} \in {\rm I\!R}^{D=3}$, the spectral densities written above take the form
%
\begin{align*}
S_{\infty}(\omega)&= \sigma^2 \sqrt{2\pi}^{D=3}  \prod_{i=1}^{D=3} \ell_i  \mathrm{exp}\left(-\frac{1}{2} \sum_{i=1}^{D=3} \ell_i^2 s_i^2 \right),   \\
S_{\frac{3}{2}}(\omega)&= \sigma^2 32\pi\sqrt{3}^3\prod_{i=1}^{D=3}\ell_i\left(3+\sum_{i=1}^{D=3}\ell_i^2 s_i^2 \right)^{-3}.
\end{align*}

@@

If we truncate the sum in (\ref{approxcov}) to the first $m$ terms, the approximate covariance function becomes
%
\begin{equation}
k(x,x') \approx \sum_{j=1}^m S_{\theta}(\sqrt{\lambda_j}) \phi_j(x) \phi_j(x') = \bm{\phi}(x)^\intercal \Delta \bm{\phi}(x'), \nonumber
\end{equation}

\noindent where $\bm{\phi}(x)=\{\phi_j(x)\}_{j=1}^{m} \in {\rm I\!R}^{m}$ is the column vector of basis functions, and $\Delta  \in {\rm I\!R}^{m\times m}$ is the diagonal matrix of the spectral densities $S_{\theta}(\sqrt{\lambda_j})$: 
%
\begin{eqnarray}
\Delta &=&  \begin{bmatrix}
    S_{\theta}(\sqrt{\lambda_1}) & & \\
    & \ddots & \nonumber \\
    & & S_{\theta}(\sqrt{\lambda_m}) \\
  \end{bmatrix}.
\end{eqnarray}

@@

\noindent As a result, the model for $f$ can be written as
%
\begin{equation}
\bm{f} \sim \mathcal{N}(\bm{\mu},\Phi \Delta \Phi^\intercal). \nonumber
\end{equation}

@@

Each multivariate eigenfunction $\phi^{\ast}_j$ corresponds to the product of the univariate eigenfunctions whose indices corresponds to the elements of the $D$-tuple $\mathbb{S}_{j\cdotp}$, and each multivariate eigenvalue $\bm{\lambda}^{\ast}_j$ is a $D$-vector with elements that are the univariate eigenvalues whose indices correspond to the elements of the $D$-tuple $\mathbb{S}_{j\bm{\cdotp}}$. Thus, for $\bm{x}=\{x_d\}_{d=1}^D \in \Omega$ and $j=1,\ldots,m^{\ast}$, we have 
%
\begin{align}
\phi^{\ast}_j(\bm{x}) &= \prod_{d=1}^{D} \phi_{\mathbb{S}_{jd}}(x_d) = \prod_{d=1}^{D} \sqrt{\frac{1}{L_d}} \text{sin}\left(\sqrt{\lambda_{\mathbb{S}_{jd}}}(x_d+L_d)\right) \label{eigenvalue_multi} \\
%
\bm{\lambda}^{\ast}_j &= \left\{ \lambda_{\mathbb{S}_{jd}} \right\}_{d=1}^D =  \left\{ \left(\tfrac{\pi \mathbb{S}_{jd}}{2L_d}\right)^2 \right\}_{d=1}^D. \label{eigenfunction_multi}
\end{align}

\noindent The approximate covariance function is then represented as
%
\begin{equation}\label{approxcov_multi}
k(\bm{x},\bm{x}') \approx \sum_{j=1}^{m^{\ast}} 
S^{\ast}_{\theta}\left(\sqrt{\bm{\lambda}^{\ast}_j}\right)
\phi^{\ast}_j(\bm{x}) \phi^{\ast}_j(\bm{x}'),
\end{equation}

\noindent where $S^{\ast}_{\theta}$ is the spectral density of the $D$-dimensional covariance function (see Section 2.1). We can now write the approximate series expansion of the multivariate function $f$ as,
%
\begin{equation}\label{approxf_multi}
f(\bm{x}) \approx \sum_{j=1}^{m^{\ast}} 
\left( S^{\ast}_{\theta} \left(\sqrt{\bm{\lambda}^{\ast}_j} \right)\right)^{1/2}
\phi^{\ast}_j(\bm{x}) \beta_j, 
\end{equation}

@@

This example consists of a simulated dataset with 250 ($i=1,\dots,n=250$) single draws of a Gaussian process prior with a Mat{\'e}rn($\nu$=3/2) covariance function and hyperparameters marginal variance $\alpha=1$ and lengthscale $\ell=0.15$, with corresponding inputs values $\bm{x}=\{x_i\}_{i=1}^{250}$ with $x_i \in [-1,1] \subset {\rm I\!R}$. To form the final noisy dataset $\bm{y}$, Gaussian noise $\sigma=0.2$ was added to the GP draws.
